Computing with an Angry Poker Player

### 2014-09-25 draft0.tex
added more sections, subsections, added written tags, started new section on acyclicity of infinite boards

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[margin=0.9in]{geometry}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{subfigure}
%\usepackage{caption}
%\usepackage{subcaption}
%\usepackage{cite}
\usepackage{hyperref}
\usepackage{url}
\urlstyle{same}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

Written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.

@@

\vbox{
\noindent
\begin{quote}
Each square of an infinite chess board contains some nonnegative number of poker chips. An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions. 
\end{quote}
}
We also consider extensions of this problem set up where the earthquake hits all squares on the board at once. We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$.
\section{Standard Board}

Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz

@@

\begin{definition} We define the big earthquake operator as a map from the set of all boards to the set of all boards and write $BE: \mathcal{B} \rightarrow \mathcal{B}$. 
It transforms all of the squares on the board with the following rule.
\begin{equation}
BE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
\end{equation}
where $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the small earthquake operator as a map from the set of all boards and locations to the set of all boards and write $SE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
If $f(x,y) \leq 4$, then $SE(f, (x,y)) = f$, otherwise,  
\begin{align*}
SE( f(x,y), (x,y)) = f(x,y)-4 \\
SE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
SE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
SE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
SE( f(x,y-1), (x,y)) = f(x,y-1)+1,
\end{align*}
and leaves all other squares unchanged.
\end{definition}

@@

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ big earthquakes or $M$ small earthquakes the resultant board will be stable.
\end{theorem}

@@

So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $BE^N(f) = BE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $SE^{M}(f) = SE^{M+k}(f)$

\begin{lemma}
If $BE(f) \neq f$, then $BE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. We will show that 
\[ BE(f) \neq f \Rightarrow \Phi(BE(f)) > \Phi(f) \]
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(BE(f)) - \Phi(f)$.
\[ \Phi(BE(f)) = \sum_{x,y} BE(f(x,y)) (|x|+|y|)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$. For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(BE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
\end{proof}

\begin{lemma}
If $SE(f) \neq f$, then $SE^n(f) \neq f$ for all $n > 0$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a small earthquake acts nontrivially on a board then the value of the potential function increases.

\begin{proof}
We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
and evaluate $\Phi(SE(f)) - \Phi(f)$. Suppose $SE$ effected square $(x_0,y_0)$ and its neighbors, then
\begin{align}
\Phi(SE(f)) - \Phi(f) =\sum_{x,y} SE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
= SE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} SE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
\end{align}
And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
\begin{align}
4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
\end{align}
\end{proof}

\begin{lemma}
\label{finiteextension}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$. Then $BE^n(f)$ and $SE^n(f)$ only contain chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
\end{lemma}

@@

@@

\section{$n$-Trees}
Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.

@@

\begin{definition}
\label{bigearthquakedeftree}
We define the big earthquake operator as  as a map from the set of all $n$ trees to the set of all $n$ trees and write $BE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$. 
It transforms all of the nodes on the tree with the following rule.
\begin{equation}
BE_T(f(x,y)) = f(x,y) - (n+1)[[f(x,y) \geq n+1]] + [[f(x-k,y_0) \geq n+1]] +  \sum_{i=1}^{n} [[f(x-k,y_i) \geq n+1]] 
\end{equation} 
where $k$ is defined as 
\begin{equation}
k = \begin{cases}
1 \text{ if } x\leq 0 \\
-1 \text{ if } x \geq 1
\end{cases}
\end{equation}
and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$ and $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the small earthquake operator as a map from the set of all $n$-trees to the set of all $n$-trees and write $SE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$.
It selects one valid node $(x,y)$ with $f(x,y) \geq n+1$ and transforms it and all of its neighbors with the following rule: 
\begin{align*}
SE_T( f( x, y ) ) = f( x , y )-n-1 \\
SE_T( f( x + k, y)) = f( x + k, y )+1 \\
SE_T( f(x-k, y_i) ) = f(x-k, y_j )+1 \\
\end{align*}
where $k$ is as defined above and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$
and leaves all other squares unchanged.
\end{definition}

@@

\begin{lemma}
If $SE_T(f) \neq f$, then $SE_T^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that  if $SE$ acts on a board nontrivially then $\Phi(SE(f))$ will be strictly larger. That is  
\begin{equation}
SE(f) \neq f \Rightarrow \Phi(SE(f)) > \Phi(f)
\end{equation}

We can evaluate $\Phi(SE(f)) - \Phi(f)$.
\begin{align}
\Phi(SE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} SE_T(f)(x,y)(x)^2 - \sum_{(x,y) \in G_n} f(x,y)(x)^2 
\end{align}
Only one node call it $(x_0,y_0)$ and its neighbors are affected. Therefore the above expression simplifies to
\begin{align*}
\Phi(SE_T(f)-\Phi(f) = SE_T( f(x_0,y_0) )* (x_0)^2 + SE_T( (f(x_0+k, y_0) )* (|x_0+k|)^2 \\
+ \sum_{i} SE_T( f(x_0-k, y_i) ) * (|x_0-k|)^2 \\
= -(n+1)*(|x_0|)^2 + (|x_0+k)^2 + n (|x_0-k|)^2 \\
\geq-(n+1)*(|x_0|)^2 + (|x_0|-1)^2 + n (|x_0|+1)^2
= -nx^2-x^2+x^2-2|x|+1+n(x^2+2|x|+1) \\
= 2x(n-1)+n+1 > 0
\end{align*}
Where $(x_0-k,y_i)$ refers to the $j$th child of $(x_0,y_0)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}.
\end{proof}

\begin{lemma}
If $BE_T(f) \neq f$, then $BE_T^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that 
\begin{equation}
BE_T(f) \neq f \Rightarrow \Phi(BE_T(f)) > \Phi(f)
\end{equation}
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(BE_T(f)) - \Phi(f)$.
\begin{align}
\Phi(BE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} BE_T(f)(x,y)(|x|)^2 - \sum_{(x,y) \in G_n} f(x,y)(|x|)^2 
\end{align}
Let $A$ be the set of active locations, or locations with $f(x,y) \geq (n+1)$, which is non-empty by assumption. 
\begin{align*}
\Phi(BE_T(f)) - \Phi(f) =  \sum_{(x,y) \in A}  BE_T(f(x,y))x^2 + BE_T(f(x+k,y))(|x+k|)^2 + \\
\sum_{i} \left( BE_T(f(x-k,y_i))(|x-k|)^2 \right) -f(x,y)x^2 +f(x+k,y) (|x+k|)^2 +  \sum_{i} \left ( f(x-k,y_i) (|x-k|)^2 \right)  \\
=\sum_{(x,y) \in A} -(n+1)x^2 + (|x+k|)^2 + n(|x-k|)^2 
\end{align*}
Where $(x-k,y_i)$ refers to the $j$th child of $(x,y)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}. We note that each term in this sum is positive by our proof of the lemma for small earthquakes and so the sum itself must be positive. That is 
\begin{equation}
\Phi(BE_T(f)) - \Phi(f) > 0
\end{equation}

@@

\begin{lemma}
\label{finiteextensiontree}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are between the levels $[x_l, x_h]$. Then $BE_T^m(f)$ and $SE_T^m(f)$ only contain chips within the range $[x_l - 2S, x_h + 2S]$ for all $m$.
\end{lemma}

@@

If we take any other arrangement of $S$ chips, with $X_{min}=l$ but not all $S$ chips are on a single node. We know that $X_{min}$ for this board is bounded below by $l-2S$. In order to have an $X_{min}$ below this at some point in time it would be possible to have $S$ chips on the same node on level $l$. But this cannot occur because in order to move a chip down a level, one chip must move up a level. Therefore if $X_{min}=x_l$, both $BE_T^m(f)$ and $SE_T^m(f)$ do not contain chips below level $x_l -2S$.

An analogous argument shows that if $X_{max}=x_h$, both $BE_T^m(f)$ and $SE_T^m(f)$ do not contain chips above level $x_h+2S$.
\end{proof}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie between levels  $[x_l, x_h]$. Then by Lemma ~\ref{finiteextensiontree}, we have that for all $m$ all chips in $BE_T^m(f)$ and $SE_T^m(f)$ will lie in $R = [x_l - 2S, x_h + 2S]$

@@

\begin{definition}
We define a valid earthquake to be a small earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.

We say that a sequence of small earthquakes $SE_1, SE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
\end{definition}

\begin{theorem}
Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ small earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
\end{theorem}

@@

\begin{lemma} Let $SE_1$ and $SE_2$ be small earthquakes. If $SE_1, SE_2$ is a valid sequence and $SE_2$ is a valid earthquake then when before $SE_1$ then $SE_2, SE_1$ is also a valid sequence of earthquakes and $SE_2( SE_1 ( B ) ) = SE_1( SE_2 ( B ) )$
\end{lemma}
\begin{proof}
Suppose $SE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $SE_2$ affects the square $(x_2, y_2)$ and its neighbors. Suppose  $SE_2, SE_1$ was an invalid sequence, that is $SE_2(B)(x_1, y_1)<4$, but since $SE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$. Therefore we must have $B(x_1, y_1)=4$ and $(x_2, y_2)=(x_1,y_1)$. But then $SE_1(B)(x_1, y_1)=3$ which means $SE_1, SE_2$ would not be a valid sequence of earthquakes. Thus $SE_2, SE_1$ is a valid sequence.

>>>>

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[margin=0.9in]{geometry}
%\usepackage{amsfonts}
%\usepackage{amsthm}
\usepackage{subfigure}
%\usepackage{caption}
%\usepackage{subcaption}
%\usepackage{cite}
\usepackage{hyperref}
\usepackage{url}
\setlength{\parindent}{0cm}
\urlstyle{same}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

@@

\emph{Written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\

@@

\vbox{
\noindent
\begin{quote}
Each square of an infinite chess board contains some nonnegative number of poker chips. An earthquake hits one square that has at least four chips and redistributes them so that one chip moves in each of the four cardinal directions. Does this process reach a stable state (with fewer than four chips on every square)? Is it independent of the order of redistribution? What can be said about the steady state?
\end{quote}
}
We also consider extensions of this problem set up where the earthquake hits all squares on the board at once. We refer to the former type of earthquake as $\textbf{local earthquakes}$ and earthquakes that hit the entire board as $\textbf{global earthquakes}$. 
\section{Definitions and Notation}

\emph{Written by Erik Waingarten, edited by Fermi Ma and Perry Kleinhenz.}\\

@@

\begin{definition} We define the global earthquake operator as a map from the set of all boards to the set of all boards and write $GE: \mathcal{B} \rightarrow \mathcal{B}$. 
It transforms all of the squares on the board with the following rule.
\begin{equation}
GE(f(x,y)) = f(x,y) - 4[[f(x,y) \geq 4]] + [[f(x-1,y) \geq 4]] + [[f(x+1,y) \geq 4]] + [[f(x,y-1) \geq 4]] + [[f(x,y+1) \geq 4]],
\end{equation}
where $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the local earthquake operator as a map from the set of all boards and locations to the set of all boards and write $LE: \mathcal{B} \times (\mathbb{Z} \times \mathbb{Z}) \rightarrow \mathcal{B}$.
If $f(x,y) \leq 4$, then $LE(f, (x,y)) = f$, otherwise,  
\begin{align*}
LE( f(x,y), (x,y)) = f(x,y)-4 \\
LE( f(x+1,y), (x,y)) = f(x+1,y)+1 \\
LE( f(x-1,y), (x,y)) = f(x-1,y)+1 \\
LE( f(x,y+1), (x,y)) = f(x,y+1)+1 \\
LE( f(x,y-1), (x,y)) = f(x,y-1)+1,
\end{align*}
and leaves all other squares unchanged.
\end{definition}

@@

\section{Stability of Finite Boards}
\emph{Written by Erik Waingarten, Edited by Fermi Ma and Perry Kleinhenz.}

\begin{theorem}
\label{finitestability}
If the initial board has a finite number of chips, there exists an $N$ and $M$ such that after $N$ global earthquakes or $M$ local earthquakes the resultant board will be stable.
\end{theorem}

@@

So a restatement of the theorem is to say that for all $B \in \mathcal{B}_f$ there exists an $N$ and $M$ such that for all $m \in \mathbb{Z}_{\geq 0}$, $GE^N(f) = GE^{N+m}(f)$ and for all $k \in \mathbb{Z}_{\geq 0}$, $LE^{M}(f) = LE^{M+k}(f)$

\subsection{A Potential Function}
\begin{lemma}
If $GE(f) \neq f$, then $GE^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
Let $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ be the following function
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2 \]
Note that this is well-defined since the boards contain a finite number of chips. We will show that 
\[ GE(f) \neq f \Rightarrow \Phi(GE(f)) > \Phi(f) \]
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(GE(f)) - \Phi(f)$.
\[ \Phi(GE(f)) = \sum_{x,y} GE(f(x,y)) (|x|+|y|)^2 \]
\[ = \sum_{x,y} \left(f(x,y) + \sum_{l \in n(x,y)} [[f(t) \geq 4]] - 4[[f(x,y) \geq 4]]\right) (|x|+|y|)^2 \]
Suppose location $f(x,y) \geq 4$, then the number of chips in location $(x,y)$ will have decreased after applying $E$. For each decrease in the number of chips atlocation $(x,y)$ of $4$, there is an increase in its four neighbors neighbors by $1$.

Let $A$ be the set of active locations, or locations which $f(x,y) \geq 4$, which is non-empty by assumption. 
So
\[ \Phi(GE(f)) - \Phi(f) \geq \sum_{(x,y) \in A} (-4)(|x|+|y|)^2 + 2(|x|+|y|+1)^2 + 2(|x|+|y|-1)^2 \]
\[ \geq \sum_{(x,y) \in A} \begin{array}{cccccc} -4x^2 & -4y^2 & -8xy & & &\\
									   +2x^2 & +2y^2 & +4xy & +2x& +2y &+ 2\\
									   +2x^2 & +2y^2 & +4xy & -2x& -2y  & +2 \end{array} \]
\[ \geq 4 > 0 \] 
\end{proof}

\begin{lemma}
If $LE(f) \neq f$, then $LE^n(f) \neq f$ for all $n > 0$.
\end{lemma}
Our proof of this lemma is essentiall the same as the preceeding one. We construct a potential function and show that if a local earthquake acts nontrivially on a board then the value of the potential function increases.

\begin{proof}
We use the same definition for $\Phi: \mathcal{B}_f \rightarrow \mathbb{Z}_{\geq 0}$ 
\[ \Phi(f) = \sum_{x,y} f(x,y)(|x|+|y|)^2, \]
and evaluate $\Phi(LE(f)) - \Phi(f)$. Suppose $LE$ effected square $(x_0,y_0)$ and its neighbors, then
\begin{align}
\Phi(LE(f)) - \Phi(f) =\sum_{x,y} LE(f(x,y))(|x|+|y|)^2 - \sum_{x,y} f(x,y)(|x|+|y|)^2 \\
= LE(f(x_0,y_0))(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} LE(f(x,y))(|x|+|y|)^2 \\- f(x_0,y_0)(|x_0|+|y_0|)^2 + \sum_{(x,y) \in n(x_0,y_0)} f(x,y)(|x|+|y|)^2 \\
= 4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2
\end{align}
And because at least two of the neighbors of $(x_0, y_0)$ are one unit farther from the origin than $(x_0, y_0)$ and at most two of the neighbors are one unit closer, and $(|x_0|+|y_0| +1)^2 > (|x_0|+|y_0| - 1)^2$, we can see,
\begin{align}
4(|x_0| + |y_0|)^2 +\sum_{(x,y) \in n(x_0, y_0)} (|x| + |y|)^2  \\ 
\geq 4(|x_0| + |y_0|)^2 + 2 (|x_0| + |y_0|+1)^2 + 2 (|x_0| + |y_0|-1)^2 = 4 >0
\end{align}
\end{proof}

\subsection{A Size Bound}
\begin{lemma}
\label{finiteextension}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are within the range $[x_l, x_h] \times [y_l, y_h]$. Then $GE^n(f)$ and $LE^n(f)$ only contain chips within the range $[x_l - \sqrt{S}, x_h + \sqrt{S}] \times [y_l-\sqrt{S}, y_h+\sqrt{S}]$ for all $n$.
\end{lemma}

@@

\subsection{Stability Result}
So now we are ready to prove the theorem that all finite boards will reach a stable state.

@@

\section{Acyclicity of Infinite Boards}

\emph{written by Fermi Ma, edited by Erik Waingarten and Perry Kleinhenz.}\\

It turns out that a potential function argument can also show that infinite boards with a bounded number of chips on each square can never cycle.

\section{Board Variants: $n$-Trees}
\emph{written by Perry Kleinhenz, edited by Fermi Ma and Erik Waingarten.}\\

Consider a semi-infinite $n$-tree. That is the graph with a parent node which has $n$ unique children. Those children each have $n$ unique children and so on.

@@

\begin{definition}
\label{bigearthquakedeftree}
We define the global earthquake operator as  as a map from the set of all $n$ trees to the set of all $n$ trees and write $GE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$. 
It transforms all of the nodes on the tree with the following rule.
\begin{equation}
GE_T(f(x,y)) = f(x,y) - (n+1)[[f(x,y) \geq n+1]] + [[f(x-k,y_0) \geq n+1]] +  \sum_{i=1}^{n} [[f(x-k,y_i) \geq n+1]] 
\end{equation} 
where $k$ is defined as 
\begin{equation}
k = \begin{cases}
1 \text{ if } x\leq 0 \\
-1 \text{ if } x \geq 1
\end{cases}
\end{equation}
and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$ and $[[A]]$ is defined as 
\begin{equation}
[[A]] = 
\begin{cases} 
1 \text{ if } A \text{ is true} \\ 
0 \text{ if } A  \text{ is false}
 \end{cases}.
\end{equation}
\end{definition}

\begin{definition}
 We define the local earthquake operator as a map from the set of all $n$-trees to the set of all $n$-trees and write $LE_T: \mathcal{T}_n \rightarrow \mathcal{T}_n$.
It selects one valid node $(x,y)$ with $f(x,y) \geq n+1$ and transforms it and all of its neighbors with the following rule: 
\begin{align*}
LE_T( f( x, y ) ) = f( x , y )-n-1 \\
LE_T( f( x + k, y)) = f( x + k, y )+1 \\
LE_T( f(x-k, y_i) ) = f(x-k, y_j )+1 \\
\end{align*}
where $k$ is as defined above and $(x-k, y_j)$ refers to the $j$th child of $(x,y)$
and leaves all other squares unchanged.
\end{definition}

@@

\begin{lemma}
If $LE_T(f) \neq f$, then $LE_T^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that  if $LE$ acts on a board nontrivially then $\Phi(LE(f))$ will be strictly larger. That is  
\begin{equation}
LE(f) \neq f \Rightarrow \Phi(LE(f)) > \Phi(f)
\end{equation}

We can evaluate $\Phi(LE(f)) - \Phi(f)$.
\begin{align}
\Phi(LE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} LE_T(f)(x,y)(x)^2 - \sum_{(x,y) \in G_n} f(x,y)(x)^2 
\end{align}
Only one node call it $(x_0,y_0)$ and its neighbors are affected. Therefore the above expression simplifies to
\begin{align*}
\Phi(LE_T(f)-\Phi(f) = LE_T( f(x_0,y_0) )* (x_0)^2 + LE_T( (f(x_0+k, y_0) )* (|x_0+k|)^2 \\
+ \sum_{i} LE_T( f(x_0-k, y_i) ) * (|x_0-k|)^2 \\
= -(n+1)*(|x_0|)^2 + (|x_0+k)^2 + n (|x_0-k|)^2 \\
\geq-(n+1)*(|x_0|)^2 + (|x_0|-1)^2 + n (|x_0|+1)^2
= -nx^2-x^2+x^2-2|x|+1+n(x^2+2|x|+1) \\
= 2x(n-1)+n+1 > 0
\end{align*}
Where $(x_0-k,y_i)$ refers to the $j$th child of $(x_0,y_0)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}.
\end{proof}

\begin{lemma}
If $GE_T(f) \neq f$, then $GE_T^n(f) \neq f$ for all $n > 0$.
\end{lemma}

\begin{proof}
We will show that 
\begin{equation}
GE_T(f) \neq f \Rightarrow \Phi(GE_T(f)) > \Phi(f)
\end{equation}
This means that if there is an earthquake that changes things, then $\Phi(f)$ increases, which means that if $E(f) \neq f$, then for all $N > 0$, $E^N(f) \neq f$, or that you never return to your original state.

We can evaluate $\Phi(GE_T(f)) - \Phi(f)$.
\begin{align}
\Phi(GE_T(f))-\Phi(f) = \sum_{(x,y) \in G_n} GE_T(f)(x,y)(|x|)^2 - \sum_{(x,y) \in G_n} f(x,y)(|x|)^2 
\end{align}
Let $A$ be the set of active locations, or locations with $f(x,y) \geq (n+1)$, which is non-empty by assumption. 
\begin{align*}
\Phi(GE_T(f)) - \Phi(f) =  \sum_{(x,y) \in A}  GE_T(f(x,y))x^2 + GE_T(f(x+k,y))(|x+k|)^2 + \\
\sum_{i} \left( GE_T(f(x-k,y_i))(|x-k|)^2 \right) -f(x,y)x^2 +f(x+k,y) (|x+k|)^2 +  \sum_{i} \left ( f(x-k,y_i) (|x-k|)^2 \right)  \\
=\sum_{(x,y) \in A} -(n+1)x^2 + (|x+k|)^2 + n(|x-k|)^2 
\end{align*}
Where $(x-k,y_i)$ refers to the $j$th child of $(x,y)$ and $k$ is as defined in Definition \ref{bigearthquakedeftree}. We note that each term in this sum is positive by our proof of the lemma for local earthquakes and so the sum itself must be positive. That is 
\begin{equation}
\Phi(GE_T(f)) - \Phi(f) > 0
\end{equation}

@@

\begin{lemma}
\label{finiteextensiontree}
Suppose $\sum_{x,y} f(x,y) = S$, and all chips are between the levels $[x_l, x_h]$. Then $GE_T^m(f)$ and $LE_T^m(f)$ only contain chips within the range $[x_l - 2S, x_h + 2S]$ for all $m$.
\end{lemma}

@@

If we take any other arrangement of $S$ chips, with $X_{min}=l$ but not all $S$ chips are on a single node. We know that $X_{min}$ for this board is bounded below by $l-2S$. In order to have an $X_{min}$ below this at some point in time it would be possible to have $S$ chips on the same node on level $l$. But this cannot occur because in order to move a chip down a level, one chip must move up a level. Therefore if $X_{min}=x_l$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not contain chips below level $x_l -2S$.

An analogous argument shows that if $X_{max}=x_h$, both $GE_T^m(f)$ and $LE_T^m(f)$ do not contain chips above level $x_h+2S$.
\end{proof}

@@

\begin{proof}
(of Theorem~\ref{finitestability}) Suppose $f \in \mathcal{B}$ is a finite board, so $\sum_{x,y} f(x,y) = S$, and suppose all chips lie between levels  $[x_l, x_h]$. Then by Lemma ~\ref{finiteextensiontree}, we have that for all $m$ all chips in $GE_T^m(f)$ and $LE_T^m(f)$ will lie in $R = [x_l - 2S, x_h + 2S]$

@@

\begin{definition}
We define a valid earthquake to be a local earthquake that occurs on a square that, at the time of the earthquake, has at least 4 chips.

We say that a sequence of local earthquakes $LE_1, LE_2, \ldots$ to be a valid sequence of earthquakes if each earthquake is valid.
\end{definition}

\begin{theorem}
Let $B$ be an initial board state that becomes a stable board $B_s$ after $N$ local earthquakes. Then for any other sequence of $N$ valid earthquakes the resultant board is identical to $B_s$.
\end{theorem}

@@

\begin{lemma} Let $LE_1$ and $LE_2$ be local earthquakes. If $LE_1, LE_2$ is a valid sequence and $LE_2$ is a valid earthquake then when before $LE_1$ then $LE_2, LE_1$ is also a valid sequence of earthquakes and $LE_2( LE_1 ( B ) ) = LE_1( LE_2 ( B ) )$
\end{lemma}
\begin{proof}
Suppose $LE_1$ affects the square $(x_1, y_1)$ and its neighbors, and $LE_2$ affects the square $(x_2, y_2)$ and its neighbors. Suppose  $LE_2, LE_1$ was an invalid sequence, that is $LE_2(B)(x_1, y_1)<4$, but since $LE_1$ is a valid earthquake we know $B(x_1, y_1)\geq 4$. Therefore we must have $B(x_1, y_1)=4$ and $(x_2, y_2)=(x_1,y_1)$. But then $LE_1(B)(x_1, y_1)=3$ which means $LE_1, LE_2$ would not be a valid sequence of earthquakes. Thus $LE_2, LE_1$ is a valid sequence.

Now consider $LE_2( LE_1 ( B ) )(u,v)$ for some $(u,v) \in \mathbb{Z} \times \mathbb{Z}$. Applying the definition of the local earthquake we see that 
\begin{align*}
LE_1 ( B )(u,v) =  B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)]] \\ 
LE_2 ( LE_1 (B)) (u,v) = LE_1(B)(u,v)  + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] \\
= B(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]]  \\
= B(u,v) + [[ (u,v) \in n(x_2, y_2) ]] - 4[[ (u,v) = (x_2, y_2)]] + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
= LE_2(B)(u,v) + [[ (u,v) \in n(x_1, y_1) ]] - 4[[ (u,v) = (x_1, y_1)] \\
= LE_1 (LE_2 (B))(u,v) 
\end{align*}
The critical step in this proof is recognizing that because $LE_1$ and $LE_2$ are valid earthquakes regardless of their ordering the algebra above was valid.
\end{proof}

\begin{lemma}
Let $LE_1,LE_2,LE_3,\dots, LE_k$ be a sequence of earthquakes, where $LE_i$ hits the square $(x_i, y_i)$. Consider  the sequence given by moving some earthquake $LE_j$ to position $m<j$ in the sequence, such that $LE_j$ is a valid earthquake for all positions between $m$ and $j$. 
This sequence is valid and produces the same final board configuration.
\end{lemma}
\begin{proof}
It is sufficient for us to show that the first $j$ earthquakes of each sequences are valid and that the board arrangement will be the same for both sequences after $j$ earthquakes. In fact we can also ignore the first $m-1$ terms, because their order is not changed they will all be valid and the board state will be identical after $m-1$ terms for both sequences. Let us call this board state $B_{m-1}$

Therefore we must show that the sequence $LE_j, LE_m, LE_{m+1}, \ldots, LE_{j-1}$ applied to $B_{m-1}$ is valid and produces the same board as the sequence $LE_m, LE_{m+1}, \ldots, LE_{j-1}, LE_j$ applied to $B_{m-1}$.
